%%  
%%  This is the file that controls everything - maxima.tex
%%
%%
%%  Initialization Stuff
\documentclass[oneside,english]{book}
\usepackage{pslatex} %%Used pslatex here because the maxima examples 
                     %%fit properly on the page in a PDF document
\usepackage[T1]{fontenc}
\usepackage[latin1]{inputenc}
\usepackage{geometry}
\usepackage{multicol}
\geometry{verbose,letterpaper,tmargin=1in,bmargin=1in,lmargin=1in,rmargin=1in}
\usepackage{babel}
\usepackage{makeidx}
\makeindex
\usepackage{emaxima}
\maximatexsessionbaselineskip = 0pt
\usepackage{graphics}
\usepackage{rotating}
\usepackage{color}
%\usepackage{mathtime}
\usepackage{titlesec}
\usepackage{thumbpdf}

\ifx\pdfoutput\undefined
     \usepackage[backref=false,
                 colorlinks=false,
                 breaklinks=true,
                 hyperindex=false]{hyperref}
%       \usepackage[dvips]{graphicx}
%               \DeclareGraphicsExtensions{.eps,.ps,.bmp,.tif,.tiff,.tga}  
%               \graphicspath{{EPS/}}
  \else
%       \usepackage[pdftex]{graphicx}
%               \pdfcompresslevel=9
%               \DeclareGraphicsExtensions{.jpg,.pdf,.mps,.png} 
%               \graphicspath{{PDF/}}
     \usepackage[pdftex=true,
                     plainpages=false,
                 backref=section,
                 colorlinks=true,
                 breaklinks=true,
%                 hyperindex=true,
                 hyperindex=false,
                 pdfpagelabels, 
%--------- Cropping -----------------------------------------------------------
% For on-line PDF generation only.
%        pdfpagescrop={98 100 512 692},
%-------- Better Color for Links ----------------------------------------------
%        linkcolor=webgreen,
%        citecolor=webgreen,
%        filecolor=webbrown,
%------ Doc Info --------------------------------------------------------------
        pdftitle={The Maxima Book},
        pdfauthor={Paulo Ney de Souza, Richard Fateman, Joel Moses, Cliff Yapp},
        pdfsubject={Maxima},
        pdfkeywords={Maxima, Macsyma, CAS, Symbolic Computation},
        pdfproducer={pdfTeX 0.14e},
        pdfcreator={LaTeX + Hyperref},
%------ Doc View --------------------------------------------------------------
        bookmarksopen=false,
        pdfnewwindow=true,
%
%       The following option controls how the document is initially displayed
%       in viewers - the whole page visible, fit width, etc.  Here are the
%       various options:
%       Fit    fit whole page in window
%       FitH   fit whole width of page 
%       FitV   fit whole height of page 
%       FitB   fit whole "Bounding Box" page 
%       FitBH  fit whole width of "Bounding Box" of page 
%       FitBV  fit whole height of "Bounding Box" of page 
%
        pdfstartview=FitH,
%
        pdfpagemode=UseOutlines]{hyperref}
%------ Fill in deficiencies of Hyperref --------------------------------------
        \pdfinfo{ 
              /CreationDate (D:20010321000000)
              /ModDate      (D:20010321000000) }
\fi

\definecolor{webgreen}{rgb}{0,.5,0}
\definecolor{webbrown}{rgb}{.6,0,0}
\definecolor{webyellow}{rgb}{0.98,0.92,0.73}

%------------ These are the fancy Chapter and Section headings:
\renewcommand{\thechapter}{\arabic{chapter}}
\titleformat{\chapter}[display]
{\bfseries\Large}
{\filleft\MakeUppercase{\chaptername} \Huge\thechapter}
{4ex}
{\titlerule
\vspace{2ex}%
\filright}
[\vspace{2ex}%
\titlerule]

\titleformat{\section}%[block]
{\large\sffamily}
{\bfseries\thesection}{.5em}{\bfseries}

\titleformat{\subsection}
%{\titlerule
{
\vspace{.8ex}%
\normalfont\itshape\bfseries}
{\hskip1cm\thesubsection.}{.5em}{}
%--------- End of fancy headings for Chapter and Sections ...


% Name of the program:
\newcommand{\maxi}{maxima}
\newcommand{\Max}{Maxima}
\newcommand{\maxp}{m\ae{}xim\raisebox{1.0ex}{\begin{turn}{-180}e\end{turn}}}

% Some custom definitions
\newdimen\firstcol
\firstcol=.35\textwidth
\newdimen\secondcol
\secondcol=.65\textwidth

\begin{document}

\title{The Maxima Book}

\author{Paulo Ney de Souza\\
        Richard J. Fateman \\
                Joel Moses \\
                Cliff Yapp}
                

\maketitle

\subsection*{Credits}

~

Paulo Ney de Souza

Jay Belanger

Richard Fateman

Joel Moses

Cliff Yapp\\
\newpage


\tableofcontents
\newpage

\addcontentsline{toc}{part}{Preface}
 \input{preface/pref.tex}

\subsection*{Examples}
\begin{multicols}{2}
\begin{tabbing}
1. First Example~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~\= \pageref{Example1} \\
2. Quitting Maxima \>\pageref{Quitting Maxima (Example 2)}\\
3. End of Entry Characters \>\pageref{End of Entry Characters (Example 3)}\\
4. Line Labels \>\pageref{Line Labels (Example 4)}\\
5. Labeling an Example Equation \>\pageref{Labeling an Equation (Example 5)}\\
6. Evaluation Toggle \>\pageref{Evaluation Toggle (Example 6)} \\
7. Basic Use of ev Command \>\pageref{Basic Use of ev Command (Example 7)}\\
8. ev's Expand Option \>\pageref{ev's Expand Option (Example 8)}\\
9. Float Example \>\pageref{FLOAT example (Example 9)}\\
\end{tabbing}
\end{multicols}

\part{The Maxima Program and Standard Packages}

\chapter{Introduction}
  \input{introduction/introduction.tex}

\chapter{Installing}
  \input{install/install.tex}

\chapter{Available Interfaces to Maxima}
  \input{interfaces/interfaces.tex}

\chapter{The Basics - What you need to know to operate in Maxima}
  \input{basics/basics.tex}

\chapter{Trig through Calculus}

    This chapter and the next will probably split into many more - Trig,
    Algebra, Calculus, Programming, etc,etc,etc. I just don't know at
    this point. These chapters will probably largely be example based.
    Using things such as ratsimp, trigsimp, etc.
   
   \input{algcal/algcal.tex}
 
\chapter{Advanced Mathematics - ODEs and Beyond}

   \input{odes/odes.tex}

\chapter{Matrix Operations and Vectors}

    Maxima has many matrix capabilities. Unfortunately it's vector handling
    at the time of this writing is not terribly robust, and the reader
    is cautioned to use it with care.

\chapter{Introduction to Maxima's Programming Language}
  \input{programming/prog.tex}


\chapter{Graphics and Forms of Output}

  \input{output/output.tex}

\chapter{Maxims for the Maxima User}
  \input{maxims/maxims.tex}

\chapter{Help Systems and Debugging}

    Using the debugger and the online help system.

\chapter{Troubleshooting}

    Common errors, their cause and solution.

\chapter{Advanced Examples}

    These examples try to draw on everything in the manual to show you
    what can be done with Maxima in advanced usages. 
    This is NOT a good starting point for new users.

   \input{examples/examples.tex}

\part{External, Additional, and Contributed Packages}


\chapter{The Concept of Packages - Expanding Maxima's Abilities}

    This chapter will try to give some in depth information on the why
    and how of Maxima packages. It is true that many {}``standard''
    abilities of Maxima, such as vectors, are due to packages, but this
    part of the book will deal with more specialized contributed packages,
    such as elliptical integrals, special scientific packages, etc. That
    way the first part of the book can be rewritten less frequently, but
    we can still have all the information here.


\chapter{Available Packages for Maxima}

    Maybe we can do this in one chapter, maybe each package will warrant
    a chapter. Not sure. For now, I'll assume one Section per package,
    not one Chapter.

\addcontentsline{toc}{chapter}{List of Figures}
\listoffigures

%\addcontentsline{toc}{chapter}{List of Tables}
%\listoftables

\addcontentsline{toc}{chapter}{Index}
\printindex

\addcontentsline{toc}{chapter}{Bibliography}
\bibliography{maxima}
\bibliographystyle{alpha}
\nocite{*}

\end{document}
