%-*-EMaxima-*-

Here we will attempt to address universal concepts which you will 
need to know when using Maxima for a wide variety of tasks. 


\section{The Very Beginning}

All computer algebra systems have syntactical rules, i.e. a structured
language by which the user communicates his/her commands to the system.
Without being able to communicate in this language, it is impossible
to accomplish anything in such as system. So we will attempt to describe
herein the basics.


\subsection{Our first Maxima Session}

We will start by demonstrating the ultimate basics: \( + \), \( - \),
\( * \), and /. These symbols are virtually universal in any mathematical
system, and mean exactly what you think they mean. We will demonstrate
this, and at the same time introduce you to your first session in
Maxima. In the interface of your choice, try the following:

\vspace{3ex}

\texttt{\label{Example1}GCL (GNU Common Lisp)~ Version(2.3.8) Wed
Sep~ 5 08:00:22 CDT 2001}

\texttt{Licensed under GNU Library General Public License}

\texttt{Contains Enhancements by W. Schelter}

\texttt{Maxima 5.5 Wed Sep 5 07:59:43 CDT 2001 (with enhancements
by W. Schelter).}

\texttt{Licensed under the GNU Public License (see file COPYING)}

\beginmaximasession
2+2;
3-1;
3*4;
9/3;
9/4;
quit();
\maximatexsession
\C1.  2+2; \\
\D1.   4 \\
\C2.  3-1; \\
\D2.   2 \\
\C3.  3*4; \\
\D3.   12 \\
\C4.  9/3; \\
\D4.   3 \\
\C5.  9/4; \\
\D5.   {{9}\over{4}} \\
\C6.  quit(); \\
\endmaximasession

\vspace{3ex}

Above is our first example of a Maxima session. We notice already
several characteristics of a Maxima session: The startup message,
which gives the version of Maxima being used, the date of compilation, 
which is the day your executable was created, the labels in front of
each line, the semicolon at the end of each line, and the way we exit
the session. The startup message is not important to the session,
but you should take note of what version of Maxima you are using,
especially if there is a known problem in an earlier version which
might impact what you are trying to do. 


\subsubsection{Exiting\index{Exiting} \index{Quitting}\index{Debugging!Exiting}a
Maxima Session}

You want to be able to get out of what you get into. So the first
command we discuss will be the command that gets you out of Maxima,
and while we are at it we will discuss how to get out of debugging
mode. Debugging mode is quite useful for some things, and the reader
is encouraged to look to later chapters for an in-depth look at the
debugging mode, but for now we will stick to basics. As you see above,
\texttt{quit();} is the command which will exit Maxima. This is a
bit confusing for new users, but you must type that full command.
Simply typing \texttt{quit} or \texttt{exit} will not work, nor will
pressing \texttt{CTRL-C} - if you try the latter you will be dumped
into the debugging mode. If that happens, simply type \texttt{:q}
if you are running GNU Common Lisp, or \texttt{:a} if running CLISP.
(If in doubt use \texttt{:q} - most binary packages use GCL at this
time.) Here's an example of what not to do, and how to get out of
it if you do:

\vspace{3ex}

\texttt{\label{Quitting Maxima (Example 2)}(C1) quit;}

\texttt{~}

\texttt{(D1)~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
QUIT}

\texttt{(C2) exit;}

\texttt{~}

\texttt{(D2)~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
EXIT}

\texttt{(C3) }

\texttt{Correctable error: Console interrupt.}

\texttt{Signalled by MACSYMA-TOP-LEVEL.}

\texttt{If continued: Type :r to resume execution, or :q to quit to
top level.}

\texttt{Broken at SYSTEM:TERMINAL-INTERRUPT.~ Type :H for Help.}

\texttt{MAXIMA>\,{}>}

\texttt{Correctable error: Console interrupt.}

\texttt{Signalled by SYSTEM:UNIVERSAL-ERROR-HANDLER.}

\texttt{If continued: Type :r to resume execution, or :q to quit to
top level.}

\texttt{Broken at SYSTEM:TERMINAL-INTERRUPT.}

\texttt{MAXIMA>\,{}>\,{}>:q}

\texttt{~}

\texttt{(C3) quit()};

\vspace{3ex}

The first two lines show what happens if you forget the \texttt{()}
or type \texttt{exit}. Nothing major, but you won't exit Maxima. \texttt{CTRL-C}
causes a few more problems - if you actually read the message, you
will see it tells you how to handle it. In the above example, \texttt{CTRL-C}
was hit twice - that is not a proper way to exit Maxima either. Just
remember to use \texttt{:q} to exit debugging and \texttt{quit();}
to exit Maxima, and you should always be able to escape trouble.  If you find
youself trying to read a long output going by quickly on a terminal, press 
\texttt{CTRL-S} to temporarily halt the output, and \texttt{CTRL-Q} to resume.


\subsubsection{The End of Entry Character}

All expressions entered into Maxima must end with either the ; character
or the \$ character. The ; character is the standard character to
use for this purpose. \index{Hiding output}The \$ symbol, while performing
the same job of ending the line, suppresses the output of that line.
This example illustrates these properties:

\vspace{3ex}

\label{End of Entry Characters (Example 3)}

\beginmaximasession
x^5+3*x^4+2*x^3+5*x^2+4*x+7;
x^5+3*x^4+2*x^3+5*x^2+4*x+7$
D2;
x^5+3*x^4+2*x^3
+5*x^2+4*x+7;
x^5+3*x^4+2*x^3
     +5*x^2+4*x+7;
\maximatexsession
\C1.  x^5+3*x^4+2*x^3+5*x^2+4*x+7; \\
\D1.   x^{5}+3\*x^{4}+2\*x^{3}+5\*x^{2}+4\*x+7 \\
\C2.  x^5+3*x^4+2*x^3+5*x^2+4*x+7$ \\
\p   \\
\C3.  D2; \\
\D3.   x^{5}+3\*x^{4}+2\*x^{3}+5\*x^{2}+4\*x+7 \\
\C4.  x^5+3*x^4+2*x^3
+5*x^2+4*x+7; \\
\D4.   x^{5}+3\*x^{4}+2\*x^{3}+5\*x^{2}+4\*x+7 \\
\C5.  x^5+3*x^4+2*x^3
     +5*x^2+4*x+7; \\
\D5.   x^{5}+3\*x^{4}+2\*x^{3}+5\*x^{2}+4\*x+7 \\
\endmaximasession

\vspace{3ex}

In (C1), we input the expression using the ; character to end the
expression, and on the return line we see that D1 now contains that
expression. In (C2), we input an identical expression, except that
we use the \$ to end the line. (D2) is assigned the contents of (C2),
but does not visually display those contents. Just to verify that
(D2) does in fact contain what we think it contains we ask Maxima
to display it's contents on (C3) and we see that the are in fact present.
This is extremely useful if you are working on a problem which has
many steps, and some of those steps would produce long outputs you
don't need to actually see. In (C4), we input part of the expression,
press return, finish the expression, and then use the ; character.
Notice that the input did not end until we used that character and
pressed return - return by itself does nothing. You see (D4) contains
the same expression as (D1) shown above. To Maxima, the inputs are
the same. This can be useful if you are going to input a long expression
and wish to keep it straight visually, to avoid errors. You can also
input spaces without adversely affecting the formula, as shown in
(C5).


\subsubsection{The (C{*}) and (D{*}) Labels}

These labels are more than just line markers - they are actually the
names in memory of the contents of the lines. This is quite useful
for a number of tasks. Let's say you wish to apply a routine, say
a \texttt{solve} routine, to an expression for several different values.
Rather than retyping the entire expression, we can use the fact that
the line numbers act as markers to shorten our task considerably,
as in this example:

\vspace{3ex}

\label{Line Labels (Example 4)}

\beginmaximasession
3*x^2+7*x+5;
solve(D1=3,x);
solve(D1=7,x);
solve(D1=a,x);
\maximatexsession
\C1.  3*x^2+7*x+5; \\
\D1.   3\*x^{2}+7\*x+5 \\
\C2.  solve(D1=3,x); \\
\D2.   \left[ x=-{{1}\over{3}},\linebreak[0]x=-2 \right]  \\
\C3.  solve(D1=7,x); \\
\D3.   \left[ x=-{{\sqrt{73}+7}\over{6}},\linebreak[0]x={{\sqrt{73}-7
 }\over{6}} \right]  \\
\C4.  solve(D1=a,x); \\
\D4.   \left[ x=-{{\sqrt{12\*a-11}+7}\over{6}},\linebreak[0]x={{\sqrt{
 12\*a-11}-7}\over{6}} \right]  \\
\endmaximasession

\vspace{3ex}

In this example, we desire to solve the expression \( 3x^{2}+7x+5 \)
for \( x \) when \( 3x^{2}+7x+5=3 \), \( 3x^{2}+7x+5=7 \), and
\( 3x^{2}+7x+5=a \). (\( a \) in this case is an arbitrary constant.)
Rather than retype the equation many times, we merely enter it once,
and then use that label to set up the similar problems more easily.

\subsubsection{The (E{*}) Labels}

In some cases, particularly when a command needs to assign generated values
to variable names, the E labels will be used.  These may be treated like any
other maxima variable.  Here is an example of E label use:

\subsubsection{Custom Labels}

You do not need to settle for this method of labeling - you can define
your own expressions if you so choose, by using the : assign operator.
Let us say, for example, that we wish to solve the problem above,
but would rather call our equation FirstEquation than D1. We will
show that process here, with one deliberate error for illustration
of a property:

\vspace{3ex}

\label{Labeling an Equation (Example 5)}

\beginmaximasession
FirstEquation:3*x^2+7*x+5;
solve(FirstEquation=3,x);
solve(FirstEquation=7,x);
solve(FirstEquation=a,x);
solve(firstequation=a,x);
\maximatexsession
\C5.  FirstEquation:3*x^2+7*x+5; \\
\D5.   3\*x^{2}+7\*x+5 \\
\C6.  solve(FirstEquation=3,x); \\
\D6.   \left[ x=-{{1}\over{3}},\linebreak[0]x=-2 \right]  \\
\C7.  solve(FirstEquation=7,x); \\
\D7.   \left[ x=-{{\sqrt{73}+7}\over{6}},\linebreak[0]x={{\sqrt{73}-7
 }\over{6}} \right]  \\
\C8.  solve(FirstEquation=a,x); \\
\D8.   \left[ x=-{{\sqrt{12\*a-11}+7}\over{6}},\linebreak[0]x={{\sqrt{
 12\*a-11}-7}\over{6}} \right]  \\
\C9.  solve(firstequation=a,x); \\
\D9.   \left[  \right]  \\
\endmaximasession

\vspace{3ex}

You see that this process works exactly the same as before. On line
(C5), you see we entered the name of FirstEquation as lower case,
and the calculation failed. These names are case sensitive. This is
true of all variables. Later on you will see cases in Maxima such
as sin, where SIN and sin are the same, but it is not safe to assume
this is always true and when in doubt, watch your cases. In general 
we suggest you use lower case for your maxima commands and programs - 
it will make them easier to read and debug.


\subsection{To Evaluate or Not to Evaluate}

Operators in Maxima, such as diff for derivative, are a common feature
in many Maxima expressions. The problem is, while you need to include
an operator at a given point in your process, you may not want to
deal with the output from it at that point in the problem. Therefore,
Maxima provides the ' toggle for operators. See the example below
for an example of how this works.

\vspace{3ex}

\label{Evaluation Toggle (Example 6)}

\beginmaximasession
diff(1/sqrt(1+x^3),x);
'diff(1/sqrt(1+x^3),x);
\maximatexsession
\C10.  diff(1/sqrt(1+x^3),x); \\
\D10.   -{{3\*x^{2}}\over{2\*\left(x^{3}+1\right)^{{{3}\over{2}}}}} \\
\C11.  'diff(1/sqrt(1+x^3),x); \\
\D11.   {{d}\over{d\*x}}\*{{1}\over{\sqrt{x^{3}+1}}} \\
\endmaximasession

\vspace{3ex}

\subsection{The Concept of Environment - The \texttt{ev} Command}

All mathematical operations in Maxima take place in an environment,
which is to say the system is assuming it should do some things and
not do other things. There will be many times you will want to change
this behavior, without doing so on a global scale. Maxima provides
a way to define a local environment on a per command basis, using
the \texttt{ev} command. \texttt{ev} is one of the most powerful commands
in Maxima, and the user will benefit greatly if they master this command
early on while using Maxima. 


\subsubsection{From the top}

We will begin with a very simple example:

\vspace{3ex}

\label{Basic Use of ev Command (Example 7)}

\beginmaximasession
ev(solve(a*x^2+b*x+c=d,x),a=3,b=4,c=5,d=6);
a;
\maximatexsession
\C1.  ev(solve(a*x^2+b*x+c=d,x),a=3,b=4,c=5,d=6); \\
\D1.   \left[ x=-{{\sqrt{7}+2}\over{3}},\linebreak[0]x={{\sqrt{7}-2
 }\over{3}} \right]  \\
\C2.  a; \\
\D2.   a \\
\endmaximasession

\vspace{3ex}

The first line uses the ev command to solve for \( x \) without setting
variables in the global environment. To make sure that our variables
remain undefined, we check that \( a \) is still undefined in line
(C2), and it is.

Now lets examine some of the more interesting features of ev. The
general syntax of the ev command is ev(exp, arg1, ..., argn). exp is
an expression, like the one in the example above. You can also use
a D{*} entry name or your own name for an expression. arg{*} has many
possibilities, and we will try to step through them here.


\subsubsection{EXPAND(m,n)}

Expand is an argument which allows you to limit how Maxima expands
an expression - i.e., how high a power you want it to expand. m is
the maximum positive power to expand, and n is the largest negative
power to expand. Here is an example:

\vspace{3ex}


\label{ev's Expand Option (Example 8)}

\beginmaximasession
ev((x+y)^5+(x+y)^4+(x+y)^3+(x+y)^2+(x+y)+(x+y)^-1+(x+y)^-2+(x+y)^-3+(x+y)^-4+(x+y)^-5,EXPAND(3,3));
\maximasession
(C3) ev((x+y)^5+(x+y)^4+(x+y)^3+(x+y)^2+(x+y)+(x+y)^-1+(x+y)^-2+(x+y)^-3+(x+y)^-4+(x+y)^-5,EXPAND(3,3));

                 1                      1                 5          4     1
(D3) ------------------------- + --------------- + (y + x)  + (y + x)  + -----
      3        2      2      3    2            2                         y + x
     y  + 3 x y  + 3 x  y + x    y  + 2 x y + x

         1          1        3        2    2      2                  3    2
    + -------- + -------- + y  + 3 x y  + y  + 3 x  y + 2 x y + y + x  + x  + x
             4          5
      (y + x)    (y + x)
\endmaximasession

\vspace{3ex}

This may be a little hard to read at first, but if you look closely
you will see that every power of \( -3\leq p\leq 3 \) has been expanded,
otherwise the subexpression has remained in it's original form. This
is extremely useful if you want to avoid filling up your screen with
large expansions that no one can read or use.


\subsubsection{Numerical Output - FLOAT and NUMER}

When one of the arguments of \texttt{ev} is FLOAT, \texttt{ev} will
convert non-integer rational numbers to floating point. NUMER will
do everything that FLOAT will, since FLOAT in invoked as part of NUMER.
NUMER also handles variables defined by the user with the NUMERVAL command,
which the FLOAT toggle will leave unevaluated.  In order to evaluate these
expressions, you can also use the \texttt{float} command.

\vspace{3ex}

\label{FLOAT/NUMER example (Example 9)}

\beginmaximasession
a:9/4;
exp(a);
ev(exp(a),FLOAT);
ev(exp(a*x),FLOAT);
numerval(b, 25);
a*b;
ev(a*b,FLOAT);
ev(a*b,NUMER);
float(a);
float(b);
float(a*b);
\maximatexsession
\C1.  a:9/4; \\
\D1.   {{9}\over{4}} \\
\C2.  exp(a); \\
\D2.   e^{{{9}\over{4}}} \\
\C3.  ev(exp(a),FLOAT); \\
\D3.   9.487735836358526 \\
\C4.  ev(exp(a*x),FLOAT); \\
\D4.   e^{2.25\*x} \\
\C5.  numerval(b, 25); \\
\D5.   \left[ b \right]  \\
\C6.  a*b; \\
\D6.   {{9\*b}\over{4}} \\
\C7.  ev(a*b,FLOAT); \\
\D7.   2.25\*b \\
\C8.  ev(a*b,NUMER); \\
\D8.   56.25 \\
\C9.  float(a); \\
\D9.   2.25 \\
\C10.  float(b); \\
\D10.   25 \\
\C11.  float(a*b); \\
\D11.   56.25 \\
\endmaximasession

\vspace{3ex}

\subsubsection{Specifying Local Values for Variables, Functions, etc.}

One of the best things about the ev command is that for one evaluation
you may specify in an arg what values are to be used for the evaluation
in place of variables, how to define functions, which functions to
evaluate, etc. We will work through a series of examples here, probably
this will be the best way to illustrate the various possibilities
of this aspect of ev.

\vspace{3ex}

\beginmaximasession
eqn1:'diff(x/(x+y)+y/(y+z)+z/(z+x),x);
ev(eqn1,diff);
ev(eqn1,y=x+z);
ev(eqn1,y=x+z,diff);
\maximatexsession
\C8.  eqn1:'diff(x/(x+y)+y/(y+z)+z/(z+x),x); \\
\D8.   {{d}\over{d\*x}}\*\left({{y}\over{z+y}}+{{z}\over{z+x}}+{{x
 }\over{y+x}}\right) \\
\C9.  ev(eqn1,diff); \\
\D9.   -{{z}\over{\left(z+x\right)^{2}}}+{{1}\over{y+x}}-{{x}\over{
 \left(y+x\right)^{2}}} \\
\C10.  ev(eqn1,y=x+z); \\
\D10.   {{d}\over{d\*x}}\*\left({{z+x}\over{2\*z+x}}+{{x}\over{z+2\*x}}
 +{{z}\over{z+x}}\right) \\
\C11.  ev(eqn1,y=x+z,diff); \\
\D11.   {{1}\over{2\*z+x}}-{{z+x}\over{\left(2\*z+x\right)^{2}}}+{{1
 }\over{z+2\*x}}-{{2\*x}\over{\left(z+2\*x\right)^{2}}}-{{z}\over{
 \left(z+x\right)^{2}}} \\
\endmaximasession

\vspace{3ex}

In this example, we define eqn1 to be the derivative of a function,
but use the ' character in front of the diff operator to notify Maxima
that we don't want it to evaluate that derivative at this time. (More
on that in the ?? section.) In the next line, we use the ev with the
diff argument, which instructs ev to take all derivatives in this
expression. Now, let's say we want to define \( y \) as a function
of \( z \) and \( x \), but again avoid evaluating the derivative.
We supply our definition of \( y \) as an argument to ev, and in
(D3) we see that the substitution has been made. Now, let's evaluate
the derivative after the substitution has been made. We work as before,
except this time we supply both the new definition of \( y \) and
the diff argument, telling ev to make the substitution and then take
the derivative. In this particular case, the order of the arguments
does not matter. The case where it will matter is if you are making
multiple substitutions - then they are handled in sequence from left
to right. 

\vspace{3ex}

(need example here, one where the difference is noticeable).

\vspace{3ex}

We can also locally define functions:

\vspace{3ex}

\beginmaximasession
eqn4:f(x,y)*'diff(g(x,y),x);
ev(eqn4,f(x,y)=x+y,g(x,y)=x^2+y^2);
ev(eqn4,f(x,y)=x+y,g(x,y)=x^2+y^2,DIFF);
\maximatexsession
\C12.  eqn4:f(x,y)*'diff(g(x,y),x); \\
\D12.   f\left(x,\linebreak[0]y\right)\*\left({{d}\over{d\*x}}\*g\left(
 x,\linebreak[0]y\right)\right) \\
\C13.  ev(eqn4,f(x,y)=x+y,g(x,y)=x^2+y^2); \\
\D13.   \left(y+x\right)\*\left({{d}\over{d\*x}}\*\left(y^{2}+x^{2}
 \right)\right) \\
\C14.  ev(eqn4,f(x,y)=x+y,g(x,y)=x^2+y^2,DIFF); \\
\D14.   2\*x\*\left(y+x\right) \\
\endmaximasession

\vspace{3ex}

(At the moment, ev seems to take only the first argument in the following
example from solve: the manual seems to indicate it should be taking
both as a list??)

\vspace{3ex}

\beginmaximasession
eqn1:f(x,y)*'diff(g(x,y),x);
eqn2:3*y^2+5*y+7;
ev(eqn1,g(x,y)=x^2+y^2,f(x,y)=5*x+y^3,solve(eqn2=5,y));
ev(eqn1,g(x,y)=x^2+y^2,f(x,y)=5*x+y^3,solve(eqn2=1,y),diff);
ev(eqn1,g(x,y)=x^2+y^2,f(x,y)=5*x+y^3,solve(eqn2=1,y),diff,FLOAT);
\maximatexsession
\C15.  eqn1:f(x,y)*'diff(g(x,y),x); \\
\D15.   f\left(x,\linebreak[0]y\right)\*\left({{d}\over{d\*x}}\*g\left(
 x,\linebreak[0]y\right)\right) \\
\C16.  eqn2:3*y^2+5*y+7; \\
\D16.   3\*y^{2}+5\*y+7 \\
\C17.  ev(eqn1,g(x,y)=x^2+y^2,f(x,y)=5*x+y^3,solve(eqn2=5,y)); \\
\D17.   \left(5\*x-{{8}\over{27}}\right)\*\left({{d}\over{d\*x}}\*
 \left(x^{2}+{{4}\over{9}}\right)\right) \\
\C18.  ev(eqn1,g(x,y)=x^2+y^2,f(x,y)=5*x+y^3,solve(eqn2=1,y),diff); \\
\D18.   2\*x\*\left(5\*x-{{\left(\sqrt{47}\*i+5\right)^{3}}\over{216}}
 \right) \\
\C19.  ev(eqn1,g(x,y)=x^2+y^2,f(x,y)=5*x+y^3,solve(eqn2=1,y),diff,FLOAT); \\
\D19.   2\*x\*\left(5\*x-0.00462962962963\*\left(\sqrt{47}\*i+5\right)
 ^{3}\right) \\
\endmaximasession

\vspace{3ex}

\subsubsection{Other arguments for ev}

INFEVAL - This option leads to an "infinite evaluation" mode, where ev 
repeatedly evaluates an expression until it stops changing. To prevent a
variable, say X, from being evaluated a way in this mode, simply include 
X='X as an argument to ev. There are dangers with this command - it is 
quite possible to generate infinite evaluation loops. For example, 
ev(X,X=X+1,INFEVAL); will generate such a loop. Here is an example: (need
example where this is useful.)

\subsubsection{How ev works}

The flow of the ev command works like this:
\begin{enumerate}
\item The environment is set up by scanning the arguments.  During this 
step, a list is made of non-subscripted variables appearing on the left 
side of equations in the arguments or in the value of some arguments if 
the value is an equation.  Both subscripted variables which do not have 
associated array functions and non-subscripted variables in the 
expression exp are replaced by their global values, except for those 
appearing in the generated list.
\item If any substitutions are indicated, they are carried out.
\item The resulting expression is then re-evaluated, unless one of the 
      arguments was NO-EVAL, and simplified according to the arguments.  
      Note that any function calls in exp will be carried out AFTER the 
      variables in it are evaluated.
\item If one of the arguments was EVAL, the previous two steps are repeated.
\end{enumerate}

\subsection{Clearing values from the system - the \texttt{kill} command}

Many times you will define something in Maxima, only to want to remove 
that definition later in the computation.  The way you do this in Maxima 
is quite simple - using the \texttt{kill} command.  Here is an example:

\beginmaximasession
A:7$
A;
kill(A);
A;
\maximatexsession
\C5.  A:7$ \\
\C6.  A; \\
\D6.   7 \\
\C7.  kill(A); \\
\D7.   \mathrm{DONE} \\
\C8.  A; \\
\D8.   A \\
\endmaximasession

\texttt{kill} is used in many situations, and has many uses.  You will 
see it appear throughout this manual, in different contexts.  There are
general arguements you can use, such as \texttt{kill(all)}, which will 
essentially start you out in a new, clean environment. (Add any relevant 
general kill options here - save kill(rules) for rules section, etc.)

\section{Common Operators in Maxima}

An operator is simply something that signals a specific operation is 
to be performed. There are many, many possible operators in Maxima.  
We will address various operators for specific jobs all throughout this 
manual - this section is not comprehensive.  

\subsection{Assignment Operators}

In mathematics, we quite often want to declare functions, assign values to 
numbers, and do many similarly useful things.  Maxima has a variety of 
operators for this purpose.

\begin{enumerate}
\item [\bf{:}] The basic assignment operator.  We have already seen this 
operator in action; it is one of the most common in maxima.
\end{enumerate}

\vspace{3ex}

\beginmaximasession
A:7;
A;
\maximatexsession
\C20.  A:7; \\
\D20.   7 \\
\C21.  A; \\
\D21.   7 \\
\endmaximasession

\vspace{3ex}

\begin{enumerate}
\item [\bf{:=}] This is the operator you would use to define functions. 
 This is a common thing to do in computer algebra, so we will illustrate
both how to and how not to do this.  
\end{enumerate}

\vspace{2ex}

The right way:

\beginmaximasession
y(x):=x^2;
y(2);
\maximatexsession
\C2.  y(x):=x^2; \\
\D2.   y\left(x\right):=x^{2} \\
\C3.  y(2); \\
\D3.   4 \\
\endmaximasession

\vspace{2ex}

Several possible wrong ways:

\beginmaximasession
y:=x^2;
y=x^2;
y(2);
y(x)=x^2;
y(2);
y[x]=x^2;
y[2];
\maximatexsession
\C22.  y:=x^2; \\
\p  Improper function definition:
y
 -- an error.  Quitting.  To debug this try DEBUGMODE(TRUE);) \\
\C23.  y=x^2; \\
\D23.   y=x^{2} \\
\C24.  y(2); \\
\D24.   y\left(2\right) \\
\C25.  y(x)=x^2; \\
\D25.   y\left(x\right)=x^{2} \\
\C26.  y(2); \\
\D26.   y\left(2\right) \\
\C27.  y[x]=x^2; \\
\D27.   y_{x}=x^{2} \\
\C28.  y[2]; \\
\D28.   y_{2} \\
\C29.  y(x):=x^2; \\
\D29.   y\left(x\right):=x^{2} \\
\C30.  y(2); \\
\D30.   4 \\
\endmaximasession

\vspace{3ex}

\begin{enumerate}
\item [~] Look over the above example - it pays to
know what doesn't work.  If you recognize the error or incorrect
result you get, it will make for faster debugging.  
\end{enumerate}

\begin{enumerate}
\item [\bf{::}] This operator is related to the : operator, but does 
not function in quite the same way.  This is more what a programmer 
would refer to as a pointer.  The best way to explain is to give you 
an example of how it behaves:
\end{enumerate}

\vspace{3ex}

\beginmaximasession
A:3$
B:5;
C:'A;
C::B;
C;
A;
\maximatexsession
\C9.  A:3$ \\
\C10.  B:5; \\
\D10.   5 \\
\C11.  C:'A; \\
\D11.   A \\
\C12.  C::B; \\
\D12.   5 \\
\C13.  C; \\
\D13.   A \\
\C14.  A; \\
\D14.   5 \\
\endmaximasession

\vspace{3ex}

You see C points to A, and A is thus assigned the value of B.

\begin{enumerate}
\item [\bf{!}] This is the factorial operator.
\end{enumerate}

\vspace{3ex}

\beginmaximasession
8!;
\maximatexsession
\C4.  8!; \\
\D4.   40320 \\
\endmaximasession

\vspace{3ex}

\begin{enumerate}
\item [\bf{!!}] This is the double factorial operator.  This is 
defined in Maxima as the product of all the consecutive odd
(or even) integers from 1 (or 2) to the odd (or even) arguement.
\end{enumerate}

\vspace{3ex}

\beginmaximasession
8!!;
2*4*6*8;
\maximatexsession
\C6.  8!!; \\
\D6.   384 \\
\C7.  2*4*6*8; \\
\D7.   384 \\
\endmaximasession

\vspace{3ex}

\begin{enumerate}
\item [\bf{sqrt(x)}] This is your basic square root operator.
\end{enumerate}

\vspace{3ex}

\beginmaximasession
sqrt(x^2);
sqrt(1/2);
sqrt(9);
\maximatexsession
\C13.  sqrt(x^2); \\
\D13.   \left| x\right|  \\
\C14.  sqrt(1/2); \\
\D14.   {{1}\over{\sqrt{2}}} \\
\C15.  sqrt(9); \\
\D15.   3 \\
\endmaximasession

\vspace{3ex}

Of course, this hardly begins to describe all the operators in the 
system, but what you see here are some of the more common and useful ones. 
